Architecture, Engineering, and Construction (AEC) projects hinge on coordination across many
stakeholders and heterogeneous data formats. Despite open digital standards (OpenBIM/IFC), day-to-day
coordination still depends on manual interpretive labour: practitioners translate unstructured site
evidence (photographs, chat notes, and marked floorplan patches) into model-linked,
schema-compliant records. This manual translation is error-prone, causing information loss across
handoffs, ambiguous evidence-to-element links, delayed update of site information, and regulatory non-compliance. We propose a
neuro-symbolic \textbf{interpreter middleware} that maps on-site evidence to the digital model
automatically by separating probabilistic perception from deterministic graph retrieval. We target
the \textbf{spatial grounding} problem at its core: linking a piece of multimodal site evidence to
the single correct IFC element (its GUID) amidst extreme geometric repetition, where dozens of
near-identical elements on a floor share every queryable attribute and only relational position
separates them. The system uses a fine-tuned VLM to emit typed spatial constraints, an IFC parse
engine to build a normalized and topology-enriched project graph, and a symbolic query cascade that
returns auditable candidate pools instead of hallucinated element IDs. This project-level grounding
layer is the core contribution; a type-conditional spatial address is an enhancement module for the
repeated-element ranking gap. On the held-out synthetic benchmark, the interpreter retains the target
in the candidate pool throughout, while learned extraction alone reaches Top-1 6.7\% because many
elements remain visually and semantically identical. The enhanced spatial address reorders the same pool from Top-1
4.9\% to \textbf{78.5\%} at the oracle ceiling; this ceiling is realizable only in part, since the
recoverable signal is the filler position-slot, where one OpenCV detector lifts the addressable
filler subset ($n=35$) from a 6.6\% floor to \textbf{58.9\%} end-to-end (39.1\% on the realistic
cluttered-floor cases) and supports selective prediction (AUROC 0.80, 95\% CI $[0.58, 0.99]$ at
$n=35$; deferring the least-confident fifth raises answered-set Top-1 to \textbf{73.4\%}). The outcome is an auditable site-to-BIM
interpreter: neural models parse evidence, IFC topology constrains retrieval, and calibrated
confidence decides when the system should defer to a human reviewer.
